---
title: "Age-structured and bottom-up interactions"
author: "James Thorson"
output: rmarkdown::html_vignette
#output: rmarkdown::pdf_document
vignette: >
  %\VignetteIndexEntry{Age-structured and bottom-up interactions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
start_time = Sys.time()
# Install locally
#  devtools::install_local( R'(C:\Users\James.Thorson\Desktop\Git\ecostate)', force=TRUE )
#  devtools::install_github( 'James-Thorson-NOAA/ecostate', force=TRUE )
# Build
# setwd(R'(C:\Users\James.Thorson\Desktop\Git\ecostate)');
# devtools::build_rmd("vignettes/too_slow/age_structured_interactions.Rmd"); rmarkdown::render( "vignettes/too_slow/age_structured_interactions.Rmd", rmarkdown::pdf_document())
```

```{r setup, echo=TRUE, message=FALSE}
library(ecostate)
```

`ecostate` is an R package for fitting the mass-balance dynamics specified by EcoSim as a state-space model.  It can also incorporate age-structured dynamics for selected multi-stanza groups, and fit to age-composition and weight-at-age data for these age-structured populations.  Here, we show the format for this using data from the Gulf of Alaska.  

## Gulf of Alaska 

We first load the example data that is included with the package:

```{r, echo=TRUE, message=FALSE, fig.width=7, fig.height=7}
# Load, inspect, and attach data
data(gulf_of_alaska)
names(gulf_of_alaska)
attach(gulf_of_alaska)
```

We then have to define various inputs used in the model.

### Defining inputs

We have to specify the discretization used when integrating differential equations for biomass dynamics, and difference equations for age-structured populations:

```{r, echo=TRUE, message=FALSE, fig.width=7, fig.height=7}
# Biomass-dynamics steps
n_step = 50

# Age-structured dynamics steps
STEPS_PER_YEAR = 24
```

We also specify default values for vulnerability and unassimilated food:

```{r, echo=TRUE, message=FALSE, fig.width=7, fig.height=7}
# Constant expected recruitment (matching assessment models)
# So h = 0.999 and back-calculate SpawnX
SpawnX = c( "Walleye pollock" = 4 / (5-1/0.999), "Sablefish" = 4 / (5-1/0.999) )

# Default vulnerability as fixed or starting values
X = array( 2, 
           dim = rep(length(taxa),2),
           dimnames = list(Prey=taxa,Predator=taxa) )

# Unassimilated food
U = array( 0.2, dim=length(taxa), dimnames=list(taxa) )
```

We then specify which parameters to estimate:

```{r, echo=TRUE, message=FALSE, fig.width=7, fig.height=7}
# Estimate catchability coefficient for all surveys
fit_Q = c( "Walleye pollock adult", "Sablefish adult", 
           "Euphausiids", "Large copepods" )

# Fit biomass-dynamics process errors for zooplankton
fit_eps = c( "Euphausiids", "Large copepods" )   #

# Fit recruitment deviations for age-structured populations
fit_phi = c("Walleye pollock", "Sablefish")

# Fit equilibrium biomass for age-structured populations
# using fishery as depletion experiment to identify scale
fit_B = c("Walleye pollock juv", "Sablefish adult")

# Fit PB with a prior
fit_PB = c( "Walleye pollock adult", "Sablefish adult" )

# Fit vulnerability for adult age-structured populations with a prior
fit_X = c("Walleye pollock adult", "Sablefish adult")
```

Finally, we define priors using a function that is applied to parameters

```{r, echo=TRUE, message=FALSE, fig.width=7, fig.height=7}
log_prior = function(p){
  lpp = 0
  # Normal(0,0.5) prior on diag(Xprime_ij), where X = exp(Xprime) + 1
  lpp = lpp + sum(dnorm( diag(p$Xprime_ij), mean=0, sd=0.5, log=TRUE )) 
  # Normal prior on log(q) for adult pollock, matching stock assessment
  lpp = lpp + dnorm( p$logq_i[match('Walleye pollock adult',taxa)], 
                     mean=log(0.85), sd=0.1, log=TRUE )
  # Tight normal prior on log(PB) for sablefish, matching assessment M value
  lpp = lpp + dnorm( p$logPB_i['Sablefish adult'], 
                     mean=log(0.1), sd=0.1, log=TRUE )
  # Tight normal prior on log(PB) for pollock, matching assessment M value
  lpp = lpp + dnorm( p$logPB_i['Walleye pollock adult'], 
                    mean=log(0.3), sd=0.1, log=TRUE )
  return(lpp)
}
```

### Building and running the model

Finally, we will build inputs without running the model

```{r, echo=TRUE, message=FALSE, fig.width=7, fig.height=7}
out = ecostate(
  taxa = taxa,
  years = years,
  catch = catch_data,
  biomass = biomass_data,
  agecomp = agecomp_data,
  PB = P_over_B,
  QB = Q_over_B,
  DC = Diet_proportions,
  B = B,
  EE = EE,
  X = X,
  type = type,
  U = U,
  fit_B = fit_B,
  fit_Q = fit_Q,
  fit_PB = fit_PB,
  fit_eps = fit_eps,
  log_prior = log_prior,
  control = ecostate_control( 
    n_steps = n_step,   
    profile = NULL, # Penalized likelihood so use empty set
    random = NULL, # Penalized likelihood so use empty set
    nlminb_loops = 0,
    getsd = FALSE ),
  settings = stanza_settings(
    taxa = taxa,
    stanza_groups = stanza_groups,
    K = K,
    Wmat = Wmat,
    Amat = Amat,
    d = d,
    Amax = Amax,
    SpawnX = SpawnX,
    fit_phi = fit_phi,
    STEPS_PER_YEAR = STEPS_PER_YEAR,
    comp_weight = "multinom",
    Leading = Leading,
    Wmatslope = Wmatslope )
)
```

We then modify starting and fixed values as needed:

```{r, echo=TRUE, message=FALSE, fig.width=7, fig.height=7}
map = out$tmb_inputs$map
tmb_par = out$obj$env$parList()

# Estimate vulnerability for adult fish with prior
map$Xprime_ij = factor( ifelse(taxa %in% fit_X, seq_along(taxa), NA)[col(array(dim=rep(length(taxa),2)))] )

# Penalized likelihood:  Fixed SD for process errors = 1
map$logtau_i = factor(rep(NA,length(map$logtau_i)))
tmb_par$logtau_i = ifelse( is.na(tmb_par$logtau_i), NA, log(1) ) 

# Penalized likelihood:  Fixed SD for recruitment deviations = 1
map$logpsi_g2 = factor(rep(NA,length(map$logpsi_g2)))
tmb_par$logpsi_g2 = ifelse( is.na(tmb_par$logpsi_g2), NA, log(1) ) 
```

Finally, we will re-build and run the model using these manual updates

```{r, echo=TRUE, message=FALSE, fig.width=7, fig.height=7}
out = ecostate(
  taxa = taxa,
  years = years,
  catch = catch_data,
  biomass = biomass_data,
  agecomp = agecomp_data,
  PB = P_over_B,
  QB = Q_over_B,
  DC = Diet_proportions,
  B = B,
  EE = EE,
  X = X,
  type = type,
  U = U,
  fit_B = fit_B,
  fit_Q = fit_Q,
  fit_PB = fit_PB,
  fit_eps = fit_eps,
  log_prior = log_prior,
  control = ecostate_control( 
    n_steps = n_step,   
    profile = NULL, # Penalized likelihood so use empty set
    random = NULL, # Penalized likelihood so use empty set
    derived_quantities = c(), # Turn off for speed
    map = map,       # Pass map back in
    tmb_par = tmb_par,  # Pass parameters back in
    getsd = FALSE ),
  settings = stanza_settings(
    taxa = taxa,
    stanza_groups = stanza_groups,
    K = K,
    Wmat = Wmat,
    Amat = Amat,
    d = d,
    Amax = Amax,
    SpawnX = SpawnX,
    fit_phi = fit_phi,
    STEPS_PER_YEAR = STEPS_PER_YEAR,
    comp_weight = "multinom",
    Leading = Leading,
    Wmatslope = Wmatslope )
)
```

We can then visualize the estimated food web

```{r, echo=TRUE, message=FALSE, fig.width=7, fig.height=7}
library(ggplot2)
plot_foodweb( 
  out$rep$out_initial$Qe_ij,
  xtracer_i = ifelse(taxa %in% c("Small phytoplankton","Large phytoplankton"),1,0),
  B_i = out$rep$out_initial$B_i,
  type_i = type )
```

We can also visualize fits to age-composition data.  To do so, we first make a helper function to extract and plot fitted and estimated values

```{r, echo=TRUE, message=FALSE, fig.width=7, fig.height=7}
do_plot = function( stanzagroup_index ){
  # Extract matrices and make into proportion at age
  Obs = out$internal$Nobs_ta_g2[[stanzagroup_index]]
  Hat = out$rep$Nexp_ta_g2[[stanzagroup_index]]
  Obs = sweep( Obs, MARGIN=1, STATS=rowSums(Obs,na.rm=TRUE), FUN="/" )
  Hat = sweep( Hat, MARGIN=1, STATS=rowSums(Hat,na.rm=TRUE), FUN="/" )
  
  # Make long-form data frame
  DF = expand.grid( 
    "Year" = rownames(Obs),
    "Age" = colnames(Obs)
  )
  DF$Age = as.numeric(DF$Age)
  DF$Obs = as.numeric(Obs)
  DF$Hat = as.numeric(Hat)
  
  #
  ggplot(DF) +
    geom_point( aes(x=Age, y=Obs) ) +
    facet_wrap( vars(Year), ncol=5, shrink=FALSE ) +    # , labeller=hospital_labeller
    geom_line( aes(x=Age, y=Hat) ) +        # , linetype="dotted"
    scale_y_continuous(name = "Proportion at age") +
    theme_classic()
}
```

We then show this for sablefish:

```{r, echo=TRUE, message=FALSE, fig.width=7, fig.height=7}
do_plot(1)
```

and for pollock

```{r, echo=TRUE, message=FALSE, fig.width=7, fig.height=7}
do_plot(2)
```

Fits confirm that the model can track cohorts for age-structured dynamics

```{r, include = FALSE, warning=FALSE, message=FALSE}
run_time = Sys.time() - start_time
```
Runtime for this vignette: `r paste( round(unclass(run_time),2), attr(run_time, "units") )`
         